An Experimental Evaluation of Hybrid Querying on Vectors (EA\&B)

### 2025-05-01 VLDB2026/main.tex
modify

	\textit{Performance of Algorithms under Multi-Attribute Building (M vs. M).}  
	As shown in Figure \ref{fig:exp_2_1}, IVF-based methods (Puck and CAPS) significantly outperform graph-based methods. The reason for Puck's superior performance is that it filters out irrelevant results during the retrieval process rather than searching for more candidates. \textcolor{red}{CAPS performs as well as Puck, but differs by further grouping results by attributes within clusters.} In contrast, FilteredVamana performs the worst because its multi-attribute indexing forces neighbor selection to cover multiple attributes, reducing the efficiency of single-attribute queries. This results in excessive traversal of irrelevant nodes, making it slower than other methods.

@@

	\textit{\textbf{Single-Attribute Selectivity.}}
	In hybrid queries, varying levels of attribute selectivity (AS) can significantly impact computational cost and query efficiency, as the algorithm must efficiently identify data points matching specific attributes within large datasets. AS is defined as the proportion of data points sharing a given attribute value. For example, AS of 1\% indicates that only 1\% of the dataset meets the given attribute.

	To investigate this effect, we evaluate 4 AS levels (1\%, 25\%, 50\%, and 75\%) on the SIFT1M dataset \textcolor{blue}{using single-thread execution. We} compare the query performance of attribute filtering algorithms under varying AS conditions, as shown in Figure~\ref{fig:exp_5_1_1_SingleLabel_1thread}.

	Experimental results show that UNG achieves the best performance across all AS levels, especially at extremely low AS (1\%). 
	This advantage stems from its pre-filtering strategy. Pre-filtering strategy significantly reduces the ANN search space by narrowing down candidates prior to vector computation. In contrast, VBASE performs the worst at 1\% AS due to its post-filtering strategy. Post-filtering strategy first retrieves a large number of candidates and then applies filtering, incurring substantial computational overhead.

@@

			\includegraphics[width=0.95\textwidth]{figures/exp/exp_5_2_1.pdf}
			\caption{Algorithms optimized for low AS levels.}
			\label{fig:exp_5_2_1}
		\end{subfigure}

@@

			\includegraphics[width=0.92\textwidth]{figures/exp/exp_5_2_2.pdf}
			\caption{Algorithms optimized for high AS levels.}
			\label{fig:exp_5_2_2}
		\end{subfigure}
		\hspace{1mm} % 增加水平间距
		\begin{subfigure}{0.32\textwidth} % 1/3 宽度
			\centering

			\includegraphics[width=0.92\textwidth]{figures/exp/exp_5_2_3.pdf}
			\caption{Algorithms insensitive to AS level.}
			\label{fig:exp_5_2_3}
		\end{subfigure}

@@

	\textcolor{blue}{Based on the} query performance trends of different algorithms under varying AS levels, we categorize the algorithms into 3 groups:

	(1) \textit{Algorithms optimized for low AS levels.}  
	This group includes ACORN-1, Faiss+HQI\_Batch, StitchedVamana, UNG, Faiss, Puck, and CAPS, as illustrated in Figure~\ref{fig:exp_5_2_1}.

	ACORN-1 performs well in low AS scenarios because fewer nodes share the same attribute. This increases the average degree of retained nodes and improves graph connectivity. CAPS efficiently skips irrelevant subpartitions, reducing computational overhead. However, as AS increases, it must process more subpartitions and handle a larger dataset. StitchedVamana optimizes local adjacency using independent subgraphs, enabling fast localization at low AS. However, at higher AS levels, it requires broader global exploration. UNG, Faiss, and Puck all use pre-filtering strategies to shrink the search space in low AS conditions, leading to better performance.

	Interestingly, Faiss and Faiss+HQI\_Batch perform worst at the 50\% AS level instead of 75\%, which might seem unexpected. This happens because, when AS decreases from 75\% to 50\%, fewer data points are filtered out. As a result, more clusters must be searched, increasing computational cost.

	\par
	(2) \textit{Algorithms optimized for high AS levels.}  
	This group includes NHQ, FilteredVamana, VBASE, and PASE, as shown in Figure~\ref{fig:exp_5_2_2}.

	At low AS levels, NHQ struggles to locate relevant regions efficiently, leading to excessive computations on non-matching nodes. FilteredVamana also performs poorly in these scenarios. Its dynamic pruning strategy makes it difficult to balance vector distance and attribute relevance, resulting in overly restricted search paths.
	VBASE relies on post-filtering, which increases computational overhead at low AS. PASE performs poorly at extremely low AS (e.g., 1\%), with recall dropping below 0.2. Its fixed candidate set may not contain enough valid results after filtering, leading to incomplete top-$k$ results and significantly reducing recall.

	\par
	(3) \textit{Algorithms insensitive to AS level.}  
	This group includes Milvus and ACORN-\(\gamma\), as illustrated in Figure~\ref{fig:exp_5_2_3}.

	Milvus is mainly limited by system-level communication overhead, such as API latency. As a result, its query performance remains largely unaffected by AS variations. ACORN-\(\gamma\) maintains stable performance across different AS levels by using a higher index construction parameter $\gamma$. This ensures a relatively stable average node degree, even after attribute pruning, minimizing the impact of AS changes on query efficiency.

@@

	Similar to the evaluation of \textit{Attribute Filtering} algorithms, we also evaluate 3 key metrics mentioned above in \textit{Range Filtering} algorithms.

@@

	\textit{\textbf{Peak memory usage.}}
	As Figure~\ref{fig:rangeFilter_memory_mb} shows, Faiss uses the least memory on low-dimensional datasets. It stores minimal temporary data (e.g., vectors, centroids, distances) during index construction.

@@

	As Figure~\ref{fig:rangeFilter_memory_mb} shows, Faiss uses the least memory on low-dimensional datasets (Deep, YT-Audio). It stores minimal temporary data (e.g., vectors, centroids, distances) during index construction.

>>>>

	\textit{Performance of Algorithms under Multi-Attribute Building (M vs. M).}  
	As shown in Figure \ref{fig:exp_2_1}, IVF-based methods (Puck and CAPS) significantly outperform graph-based methods. The reason for Puck's superior performance is that it filters out irrelevant results during the retrieval process rather than searching for more candidates. The performance of CAPS is as good as that of Puck. The difference between them is that CAPS further groups the results by attributes within clusters. In contrast, FilteredVamana performs the worst because its multi-attribute indexing forces neighbor selection to cover multiple attributes, reducing the efficiency of single-attribute queries. This results in excessive traversal of irrelevant nodes, making it slower than other methods.

@@

	\textit{\textbf{Single-Attribute Selectivity.}}
	In hybrid queries, varying attribute selectivity (AS) can significantly impact computational cost and query efficiency, as the algorithm must efficiently identify data points matching specific attributes within large datasets. AS is defined as the proportion of data points sharing a given attribute value. For example, AS of 1\% indicates that only 1\% of the dataset meets the given attribute.

	To investigate this effect, we evaluate 4 AS \textcolor{blue}{settings} (1\%, 25\%, 50\%, and 75\%) on the SIFT1M dataset \textcolor{blue}{using single-thread execution. We} compare the query performance of attribute filtering algorithms \textcolor{blue}{across these settings}, as shown in Figure~\ref{fig:exp_5_1_1_SingleLabel_1thread}.

	Experimental results show that UNG achieves the best performance across all AS  \textcolor{blue}{settings}, especially at extremely low AS (1\%). 
	This advantage stems from its pre-filtering strategy. Pre-filtering strategy significantly reduces the ANN search space by narrowing down candidates prior to vector computation. In contrast, VBASE performs the worst at 1\% AS due to its post-filtering strategy. Post-filtering strategy first retrieves a large number of candidates and then applies filtering, incurring substantial computational overhead.

@@

			\includegraphics[width=0.95\textwidth]{figures/exp/exp_5_2_1.pdf}
			\caption{Algorithms optimized for low AS.}
			\label{fig:exp_5_2_1}
		\end{subfigure}

@@

			\includegraphics[width=0.92\textwidth]{figures/exp/exp_5_2_2.pdf}
			\caption{Algorithms optimized for high AS.}
			\label{fig:exp_5_2_2}
		\end{subfigure}
		\hspace{1mm} % 增加水平间距
		\begin{subfigure}{0.32\textwidth} % 1/3 宽度
			\centering

			\includegraphics[width=0.92\textwidth]{figures/exp/exp_5_2_3.pdf}
			\caption{Algorithms insensitive to AS.}
			\label{fig:exp_5_2_3}
		\end{subfigure}

@@

	\textcolor{blue}{Based on the} query performance trends of different algorithms under varying AS \textcolor{blue}{settings}, we categorize the algorithms into 3 groups:

	(1) \textit{Algorithms optimized for low AS.}  
	This group includes ACORN-1, Faiss+HQI\_Batch, StitchedVamana, UNG, Faiss, Puck, and CAPS, as illustrated in Figure~\ref{fig:exp_5_2_1}.

	ACORN-1 performs well in low AS \textcolor{blue}{settings} because fewer nodes share the same attribute. This increases the average degree of retained nodes and improves graph connectivity. CAPS efficiently skips irrelevant subpartitions, reducing computational overhead. However, as AS increases, it must process more subpartitions and handle a larger dataset. StitchedVamana optimizes local adjacency using independent subgraphs, enabling fast localization at low AS. However, at higher AS \textcolor{blue}{settings}, it requires broader global exploration. UNG, Faiss, and Puck all use pre-filtering strategies to shrink the search space in low AS \textcolor{blue}{settings}, leading to better performance.

	Interestingly, Faiss and Faiss+HQI\_Batch perform worst at the 50\% AS instead of 75\%.
	%, which might seem unexpected. 
	This happens because, when AS decreases from 75\% to 50\%, fewer data points are filtered out. As a result, more clusters must be searched, increasing computational cost.

	\par
	(2) \textit{Algorithms optimized for high AS.}  
	This group includes NHQ, FilteredVamana, VBASE, and PASE, as shown in Figure~\ref{fig:exp_5_2_2}.

	At low AS, NHQ struggles to locate relevant regions efficiently, leading to excessive computations on non-matching nodes. FilteredVamana also performs poorly in these scenarios. Its dynamic pruning strategy makes it difficult to balance vector distance and attribute relevance, resulting in overly restricted search paths.
	VBASE relies on post-filtering, which increases computational overhead at low AS. PASE performs poorly at extremely low AS (e.g., 1\%), with recall dropping below 0.2. Its fixed candidate set may not contain enough valid results after filtering, leading to incomplete top-$k$ results and significantly reducing recall.

	\par
	(3) \textit{Algorithms insensitive to AS.}  
	This group includes Milvus and ACORN-\(\gamma\), as illustrated in Figure~\ref{fig:exp_5_2_3}.

	Milvus is mainly limited by system-level communication overhead, such as API latency. As a result, its query performance remains largely unaffected by AS variations. ACORN-\(\gamma\) maintains stable performance across different AS \textcolor{blue}{settings} by using a higher index construction parameter $\gamma$. This ensures a relatively stable average node degree, even after attribute pruning, minimizing the impact of AS changes on query efficiency.

@@

	%Similar to the evaluation of \textit{Attribute Filtering} algorithms, 
	\textcolor{blue}{Similarly}, we also evaluate 3 key metrics mentioned above in \textit{Range Filtering} algorithms.

@@

	\textit{\textbf{Peak memory usage.}}
	As Figure~\ref{fig:rangeFilter_memory_mb} shows, Faiss uses the least memory on low-dimensional datasets. It stores minimal temporary data during index construction.

@@

	%As Figure~\ref{fig:rangeFilter_memory_mb} shows, Faiss uses the least memory on low-dimensional datasets (Deep, YT-Audio). It stores minimal temporary data (e.g., vectors, centroids, distances) during index construction.
